Template Master's Thesis

### 2023-08-11 paper.tex
Corrections and updates until clustering heuristics

A number of address clustering heuristics in the UTXO setup have been proposed for Bitcoin and derivatives \citep{Androulaki2013, Meiklejohn2013, Haslhofer2016, jourdan2018, kappos2022}. All these methods will not be the focus of this work and therefore not discussed. As suggested by \cite{nakamotoBitcoin2008} in the Bitcoin whitepaper, most Bitcoin wallet implementations use a new key pair for each transaction to keep them from being linked to a common owner. \newline
In contrast to the UTXO-model, native transactions in account-based blockchains can only move funds between a single sender and receiver, and the ``change`` remains in the sender account. Subsequent transactions necessarily use the same address again. The account-based model essentially relies on address reuse on the protocol level \citep{Beres2020}. Therefore, privacy guarantees should be lower and most users likely use only a small number of addresses. \newline
% Make an assumption that people only use a small set of addresses
Clustering heuristics for account-based blockchains were first introduced by \cite{FV:17}. He proposed and applied heuristics that exploit patterns related to deposit addresses, multiple participation in airdrops, and token authorization mechanisms. \newline
\cite{Beres2020} propose more universally applicable methods, as they argue that Victor’s heuristics, while powerful, assume participation in certain on-chain events. Their technique interprets transactions or token transfers as network graphs, with addresses as nodes and asset transfers as edges. They quantitavely compare graph-representation learning algorithms (a subset of machine learning) and propose further user profiling techniques based on time-of-day activity and transaction fees. Using ENS address pairs as ground truth, the authors rigorously test their methods and apply their findings to significantly reduce the privacy guarantees of \textit{Tornado Cash}, a non-custodial privacy-enhancing protocol on Ethereum. \newline
\cite{wu2022tutela} extend on one of \cite{Beres2020} graph representation learning %node-embedding
 algorithms and apply it at scale. Further, they also propose a set of new clustering heuristics targeting Tornado Cash,  showing that careless user behavior can still reveal identity. Based on those heuristics, they build an application built to measure the anonymity of an Ethereum address. \newline All of the methods mentioned will be discussed in greater detail in Section 3. \newline
Broadening the scope of entity identification, \cite{victorlüders2019} study the largest Ethereum ERC-20 token networks from a graph perspective. Similarly, \cite{casalebrunet2021} apply network graph analysis to various non-fungible token (NFT) ecosystems. Both find that many of the networks follow either a star or a hub-and-spoke pattern. These patterns are common to interaction graphs in social networks. Further, \cite{Payette2017} propose a segmentation of the Ethereum address space into four distinct behaviour groups sharing similar attributes while using k-means clustering. \newline
Rather than treating users as entities and analyzing on-chain data, \cite{yu2023} propose (a novel) approach for correlation analysis by exploiting network information. Although this approach has the potential to avoid the impact of privacy-enhancing technologies, it introduces a new set of limitations and problems. Thus, this approach may be of great interest for further research, particularly when privacy-enhancing techniques become more widely used.

\textbf{Our contribution}\\
In this work, we perform entity identification on an address set containing 473,927 addresses. To obtain these addresses, a separate project collected snapshots of avatar activity in Decentraland over an ten-month period from July 2022 to April 2023. Due to Decentraland's blockchain-based architecture, each avatar contains information about a user's Web3 address. \newline
We test the applicability of existing heuristics, and if feasible, their efficacy in detecting entities using multiple addresses is evaluated.
The main objective is to estimate the number of real-world entities that these addresses represent. The clustering of addresses is limited to this set only, without considering any %(clusters for)
 addresses outside of it, e.g., that were not registered in Decentraland during the given time frame. \newline
Furthermore, we introduce our own heuristics for identifying address clusters and evaluate their suitability for implementation in other contexts.
%Anwenden/discuss and evaluate existing heuristics to my specific address set.
%Anpassen, erweitern

@@

\section{Preliminaries}
% Introduce section
\subsection{Contract Accounts and EOAs}
Account-based blockchains usually distinguish between \textit{externally owned accounts} (EOAs) and \textit{contract accounts} (smart contracts). EOAs are created and controlled by private keys and can be used to hold the native protocol asset, send and receive transactions, and interact with contract accounts. Contract accounts are controlled by the contract's code, their state can be modified through transactions sent to the contract and they cannot initiate transactions. \citep{buterin2014ethereum} \newline Each account has a 20-byte address encoded in hexadecimal. For EOAs, this address is based on the last 20 bytes of the Keccak-256 hash of the ECDSA public key. For contract accounts, it is usually the last 20 bytes of the Keccak-256 hash of the RLP encoding of the sender address and account nonce. \citep{GW:14}

\subsection{Ethereum, Polygon and the EVM}
Ethereum and Polygon are both smart contract based account-model blockchains. They share the same execution logic - the Ethereum Virtual machine (EVM).
The EVM provides a standardized framework for contract execution, ensuring that smart contracts produce deterministic results across all nodes in the network \citep{GW:14}. The EVM is independent of the underlying blockchain protocol. This allows other blockchain implementations to adopt this standardized framework for contract execution.\newline
Polygon is a so-called sidechain and aims to address the scalability limitations of Ethereum. The adoption of the EVM means that both blockchains share the same user address schemes. It further enables developers to deploy and execute Ethereum-based smart contracts on the Polygon network with minimal changes. \citep{matic_whitepaper}
%A high-level programming language is typically used to write smart contracts, and a compiler to convert them into bytecode \citep{mastering_ethereum}. Solidity is currently the dominant smart contract language.
%Polygon PoS is a so-called sidechain, closely associated with Ethereum mainnet and used for scaling.
%Ethereum Mainnet is the chain on which much of the Decentraland infrastructure has been deployed.
%In addition, \cite{eip1014} introduces a new contract creation mechanism allowing for more predictable contract addresses.%EIP-1014
%This allows developers to calculate the address of a contract before it is deployed.

\subsection{Tokens and Token Standards}
\cite{roth2019tokenization} define \textit{tokens} as rivalrous, digital units of value that represent ownership of an asset or utility. Smart contracts are the primary method of creating tokens on EVM-based blockchains. In essence, a smart-contract-based token is a mapping of accounts with token balances and a set of functions defining how these balances can be changed. Any smart contract containing these elements can be interpreted as a token contract. \citep{roth2019tokenization} \newline
\textit{Token standards} specify basic interfaces that allow for interoperability between contracts. These standards do not prescribe an implementation but set minimum requirements without restricting the design beyond that \citep{mastering_ethereum}. 
The most common token standards are ERC-20\footnote{https://eips.ethereum.org/EIPS/eip-20} for fungible tokens, ERC-721\footnote{https://eips.ethereum.org/EIPS/eip-721} for non-fungible tokens (NFTs) and ERC-1155\footnote{https://eips.ethereum.org/EIPS/eip-1155} for semi-fungible tokens.\newline
When a transaction completes, it produces a transaction receipt  that contains log entries providing information about the actions that occurred during its execution. Solidity\footnote{Solidity is the dominant high-level programming language for smart contracts targeting the EVM.} high-level objects called \textit{Events} construct these logs, which can be queried from a full node and are stored separately from the state. \citep{mastering_ethereum} \newline
The standarization of tokens allows us to listen for ``transfer events``, which are emitted whenever an token changes ownership.
These transfer events include the \textit{sender} and \textit{recipient} address, along with a \textit{token ID} and/or \textit{amount}.

@@

\subsection{Decentraland}
\textit{Decentraland} is the first large-scale blockchain-based virtual world. It has attracted the attention of well-known companies like Tommy Hilfiger, Samsung, PepsiCo, Diesel, Adidas, and Netflix, who are actively engaged in it  \citep{metaverse-retailing2023}.\newline
Decentraland is an ideal platform for empirical research for two main reasons.
\textit{First}, its open architecture allows compiling snapshots of user activity, including the users' locations  within the metaverse. This was done by \cite{metaverse-retailing2023}, who captured snapshots of avatar locations over a period of 10 months. The addresses they collected are used for this thesis.
\textit{Second}, users who connect their Web3 account can own various digital assets, such as monetary units, land parcels, and avatar collectibles (e.g. wearables, emotes, and names). Smart contracts on Ethereum and Polygon track and manage the ownership of these assets \citep{goldbergschaer2023}. This allows us to analyze the entire transaction history and derive information about the users' economic activity. %avatars'
\newline
Initially, Decentraland launched with two native tokens. \textit{LAND}, an NFT compliant with ERC-721, manages the ownership of land parcels in Decentraland. \textit{MANA}, a fungible token compliant with ERC-20, is the in-world currency used to purchase digital goods and services. The majority of purchases and trades are settled in MANA. \newline 
Moreover, the Decentraland core team/developers provide a smart-contract-based marketplace where the users may buy or exchange LAND or other in-game collectibles. %the foundation/DAO/developers

@@

\subsection{Non-custodial crypto asset mixers}
%intuition of how it works,  distinction between custodial and non-custodial, Tornado cash = SC based
According to \cite{nadler2023tornado}, crypto-asset mixers are currently the most widely used approach to achieve privacy on public blockchains. At a high level, we can distinguish between custodial and non-custodial crypto-asset mixers. \newline
In the custodial setup, users send their assets to a centralized service provider's public deposit address. In return, the provider sends the assets back to a privately relayed recipient address. This approach is fully trust-based, as the service provider has complete control over the assets and access to the identifying data. \newline 
In contrast, non-custodial crypto asset mixers (e.g., Tornado Cash) replace the trusted mixing party with a publicly verifiable smart contract. They rely on cryptographic schemes that allow anyone to prove and independently verify the validity of a withdrawal without disclosing the link to a specific deposit. This has the advantage that users do not have to share the identifying information with anyone and there is no liquidity risk as the funds are locked. \citep{nadler2023tornado} \newline
In a nutshell, the mixer works in the following way: Various entities deposit the same amount of a specific crpyto asset to a mixer address acting as a pool. Anyone who has contributed to the pool may then generate a new address and withdraw their funds without revealing the link between the deposit and withdrawal addresses. This is achieved using zkSNARKs (zero knowledge, succinct, non-interactive argument of knowledge). 
Each depositor inserts a hash value in a Merkle-tree. At the time of withdrawal, each legitimate withdrawer can prove unlinkably with a zero-knowledge proof that they know the pre-image of a previously inserted hash leaf in the Merkle-tree. %Subsequently, users can withdraw their asset from the mixer whenever they consider that the size of the anonymity set is satisfactory.
 For a more detailed description of how Tornado Cash works, see \cite{nadler2023tornado} or \cite{Beres2020}. \newline %For a mathematical discussion around zkSNARKs, please refer to Petkus (2019), Thaler (2022) or Berentsen et al. (2022).
In the context of this work, it is sufficient to know that third parties can still observe the addresses that have deposited to and withdrawn from the pool. The set of all deposits that a particular withdrawal could be originated from is defined as the \textit{anonymity set}. 
In theory, a larger anonymity set will provide a higher level of privacy, since  any withdrawal address could be linked to a larger set of deposit addresses. However, in practice this does not always hold true. %this is not always the case
Some users reveal the link between their withdrawal and deposit addresses, which also affects the privacy of other depositors. For example, by withdrawing from the pool with the same address they deposited or directly send funds between the two addresses. According to \cite{nadler2023tornado}, careless use of the protocol, testing transactions or external incentive mechanisms might be reasons for this behaviour. \newline
In addition, \cite{Beres2020} and \cite{wu2022tutela} propose approaches to de-anonymize Tornado cash transactions using various clustering heuristics. These methods are discussed in more detail in section 4.

@@

\section{Data Collection and Preparation}
% Introduction Satz - Ausgangslage 
The starting point for our work was a dataset comprising 473,927 distinct (Web3) addresses. %The original dataset comprised 473,927 distinct (Web3) addresses.
 As we focus on clustering methods using on-chain data, we exclude addresses that have not been recorded on either Ethereum or Polygon. To accomplish this, we used data from Blockscan\footnote{\url{https://blockscan.com/}}. Overall, 59,651 addresses were recorded on Ethereum, 129,988 on Polygon, with 52,095 addresses appearing on both networks simultaneously. The set of remaining addresses consists of 137,544 addresses. Subsets are visualized as a Venn diagram in Figure \ref{fig:Venn}.

@@

\subsection{Transactions and Token Transfer Events}
The address subsets were used to collect transaction and token transfer event data from Ethereum and Polygon. Ethereum or Polygon data can be accessed directly through a node or an application programming interface (API) provider like Infura or Etherscan. The absence of account indexing in Ethereum or Polygon poses a challenge for retrieving all past transactions and token transfers of a specific address, as it requires scanning through all blocks and token transfer events emitted by designated token contracts. Fortunately, Etherscan offers an API Endpoint Module for ``Accounts`` that facilitates the retrieval of transactions and token transfer events for a given address. By using our network-specific address subsets, we were able to significantly decrease the number of required API calls. 
We gathered all transactions and token transfers up until block \texttt{17,670,000} on Ethereum (July 11, 2023) and block \texttt{44,990,000} on Polygon (July 12, 2023). In total, we collected more than 30 million %30,689,978
token transfers and more than 16 million %16,092,531
 normal transactions. While the amount of transactions is similar comparing between the two networks, there are 3x more transfer events on polygon. The database contains distinct collections for transactions and transfers. See appendix for information on the different fields.

The output was stored in separate CSV files and also imported into a MongoDB\footnote{\url{https://www.mongodb.com/}} database.

@@

\subsection{Intra-set asset transfers}
When it comes to entity clustering, we essentially want to answer the question of how many real-world entities our remaining  137,544 addresses represent. Another way to put this would be: Are users interacting through multiple EOAs with Decentraland and if so, are we able to detect them? (logging in to Decentraland with multiple addresses) \newline
To achieve this, we often only consider transactions or token transfers where both the 'from' and the 'to' address are within this narrowed pool of addresses. For normal transactions, this means that we only consider native asset transfers (Ether or matic). \newline
As nearly all addresses are Externally Owned Accounts (EOAs), we do not need to differentiate between EOAs and smart contract accounts, which further simplifies the clustering process. \newline
We use this dataset to generate the network graphs. Also for other clustering heuristics we often make this reduction, for example: In the self-approval approach, we only consider approval transactions where another address from the set is approved.

%filter our transactions and token transfers to entries where both the 'from' and the 'to' are within this narrowed pool of addresses. We define this as intra-set asset transfers. For the transactions, this means that we only consider native asset (Ether or matic) transfers. %between two EOAs
%As nearly all addresses are Externally Owned Accounts (EOAs), we do not need to differentiate between EOAs and smart contract accounts, which further simplifies the process.

%As already mentioned in the introduction, the goal of this paper is to cluster addresses within who (1) were active in Decentraland during the 9-month period (e.g., within the initial dataset) and (2) transacted or interacted with any token on Ethereum or Polygon. 
%Essentialy, we want to answer two main research questions when it comes to clustering:
%1. How many unique entities does the address set represent (i.e. were active on Decentraland during the 9-month period)?
%2. Are users interacting through multiple accounts/EOAs/addresses with Decentraland?
%we focus on on-chain data (and not user location or avatar data), allowing us to exclude addresses that have not transacted or interacted with any token on Ethereum or Polygon. 
%As mentioned in the introduction, the goal of this paper is to cluster addresses belongig to the same entity that were recorded/active in Decentraland during the 9-month period. Therefore, we disregard addresses outside of this set. Also, since we only look at on-chain data, we cannot cluster addresses without activity on Polygon or Ethereum. Doing these two reductions, there are  141, 591 'clusterable' addresses remaining.
%Since all of our addresses are EOAs (some multisig smart contract wallets). -> Check this?

\subsection{ENS ground-truth pairs}
Similarly to \cite{Beres2020}, we use Ethereum Name Service (ENS) identifiers as ground truth information to evaluate some of the clustering methods.\newline
\textit{ENS} is a naming system that relies on the Ethereum smart contracts. Its main purpose is to map human-readable names, such as \texttt{`alice.eth`}, to machine-readable identifiers, like Ethereum addresses. The architecture of ENS comprises two main components: the registry and resolvers. \newline
The \textit{ENS registry} consists of a single smart contract\footnote{The registry contract is ERC-721 compliant, which means that .eth registrations can be transferred in the same way as other NFTs.} that maintains a list of all domains and subdomains, recording the ``owner`` and ``resolver`` for each. The registry allows the owner of a domain to make changes to that data. %Domain owners may set the resolver for the domain.
\newline
\textit{Resolvers} are responsible for %(the actual process of)
 translating names into addresses. Any contract that adheres to the relevant standards may act as a resolver in ENS. The process of resolving a name involves two steps: \textit{First}, the registry must be queried to find out which resolver is responsible for the name, and \textit{second}, that resolver must be asked for the response to the query. \newline
In addition to regular resolution from name to address, ENS also supports ``reverse resolution``, allowing for a mapping from address back to a name (or other metadata). Reverse resolution is accomplished via the special domain \texttt{`addr.reverse`}% and the resolver function \texttt{name()}.
. This domain is owned by a special purpose registrar contract that allocates subdomains to the owner of the matching address - for instance, the address \texttt{`0x1234\dots`} may claim the name \texttt{`1234\dots.addr.reverse`}, and configure a resolver and records on it. The resolver in turn supports the \texttt{name()} function, which returns the name associated with that address.\newline
 Reverse resolution %(without a library)
 follows the same pattern as forward resolution: Get the resolver for \texttt{`1234....addr.reverse'} (where \texttt{`1234...`} is the address you want to reverse-resolve), and call the \texttt{name()} function on that resolver. However, ENS does not enforce the accuracy of reverse records. This means \texttt{`1234....addr.reverse`} may claim that the name associated to their address is \texttt{`alice.eth`}.

Figure \ref{fig:ENS} illustrates the mechanism that allowed us to generate ground truth pairs: Initially, the domain name \texttt{`foo.eth`} resolves to address \texttt{`0xABC...`}. The owner also set a reverse resolution for this address, indicated by the the edge from the address to the domain name. Next, the owner of \texttt{`foo.eth`} decides to change the resolved address to \texttt{`0xDEF...`}, again setting a reverse record.  Nevertheless, \texttt{`ABC....addr.reverse`} still points to \texttt{`foo.eth`}. If we query the name corresponding to both \texttt{`0xABC...`} and \texttt{`0xDEF...`}, both will return the same name, even though a specific name can only resolve to one address at a time. 
We make the assumption that both addresses belong to the same entity if this specific case occurs.
%We did not observe cases where someone set a reverse record to a completely unrelated address (e.g. alice.eth)

>>>>

A number of address clustering heuristics in the UTXO setup have been proposed for Bitcoin and derivatives \citep{Androulaki2013, Meiklejohn2013, Haslhofer2016, jourdan2018, kappos2022}. All these methods will not be the focus of this work and therefore not discussed. \newline 
As suggested by \cite{nakamotoBitcoin2008} in the Bitcoin whitepaper, most Bitcoin wallet implementations use a new key pair for each transaction to keep them from being linked to a common owner. In contrast to the UTXO-model, native transactions in account-based blockchains can only move funds between a single sender and receiver, and the ``change`` remains in the sender account. Subsequent transactions necessarily use the same address again. The account-based model essentially relies on address reuse on the protocol level \citep{Beres2020}. Therefore, privacy guarantees should be lower as most participants are likely to use only a small number of addresses. \newline
% Make an assumption that people only use a small set of addresses
Clustering heuristics for account-based blockchains were first introduced by \cite{FV:17}. He proposed and applied heuristics that exploit patterns related to deposit addresses, multiple participation in airdrops, and token authorization mechanisms. \newline
\cite{Beres2020} propose more universally applicable methods, as they argue that Victor’s heuristics, while powerful, assume participation in certain on-chain events. Their technique interprets transactions or token transfers as network graphs, with addresses as nodes and asset transfers as edges. They quantitavely compare graph-representation learning algorithms (a subset of machine learning) and propose further user profiling techniques based on time-of-day activity and transaction fees. Using ENS address pairs as ground truth, the authors rigorously test their methods and apply their findings to significantly reduce the privacy guarantees of \textit{Tornado Cash}, a non-custodial privacy-enhancing protocol on Ethereum. \newline
\cite{wu2022tutela} extend on one of \cite{Beres2020} graph representation learning %node-embedding
 algorithms and apply it at scale. Further, they also propose a set of new clustering heuristics targeting Tornado Cash,  showing that careless user behavior can still reveal identity. Based on those heuristics, they built an application to measure the anonymity of an Ethereum address. \newline All of the methods mentioned will be discussed in greater detail in Section 3. \newline
Broadening the scope of entity identification, \cite{victorlüders2019} study the largest Ethereum ERC-20 token networks from a graph perspective. Similarly, \cite{casalebrunet2021} apply network graph analysis to various non-fungible token (NFT) ecosystems. Both find that many of the networks follow either a star or a hub-and-spoke pattern. These patterns are common to interaction graphs in social networks. Further, \cite{Payette2017} propose a segmentation of the Ethereum address space into four distinct behaviour groups sharing similar attributes while using k-means clustering. \newline
Rather than treating users as entities and analyzing on-chain data, \cite{yu2023} propose (a novel) approach for correlation analysis by exploiting network information. Although this approach has the potential to avoid the impact of privacy-enhancing technologies, it introduces a new set of limitations and problems. Thus, this approach may be of great interest for future research, particularly when privacy-enhancing techniques become more widely used.

\textbf{Our contribution}\\
In this work, we perform entity identification on an address set containing 473,927 addresses. To obtain these addresses, a separate project collected snapshots of avatar activity in Decentraland over an nine-month period from July 2022 to April 2023. Due to Decentraland's blockchain-based architecture, each avatar contains information about a user's Web3 address. \newline
We test the applicability of existing heuristics, and if feasible, their efficacy in detecting entities using multiple addresses is evaluated.
The main objective is to estimate the number of real-world entities that these addresses represent. The clustering of addresses is limited to this set only, without considering any %(clusters for)
 addresses outside of it, e.g., that were not registered in Decentraland during the given time frame. \newline
Furthermore, we introduce our own heuristics for identifying address clusters and evaluate their suitability for implementation in other contexts.
%Anwenden/discuss and evaluate existing heuristics to my specific address set.
%Anpassen, erweitern

@@

\section{Preliminaries}
% Introduce section
\subsection{Contract Accounts and EOAs}
Account-based blockchains usually distinguish between externally owned accounts (EOAs) and contract accounts (smart contracts). \textit{EOAs} are created and controlled by private keys and can be used to hold the native protocol asset, send and receive transactions, and interact with contract accounts. \textit{Contract accounts} are controlled by the contract's code, their state can be modified through transactions sent to the contract and they cannot initiate transactions. \citep{buterin2014ethereum} \newline Each account has a 20-byte address encoded in hexadecimal. For EOAs, this address is based on the last 20 bytes of the Keccak-256 hash of the ECDSA public key. For contract accounts, it is usually the last 20 bytes of the Keccak-256 hash of the RLP encoding of the sender address and account nonce. \citep{GW:14}

\subsection{Ethereum, Polygon and the EVM}
Ethereum and Polygon are both smart contract based account-model blockchains. They share the same execution logic - the Ethereum Virtual machine (EVM).
The EVM provides a standardized framework for contract execution, ensuring that smart contracts produce deterministic results across all nodes in the network \citep{GW:14}. The EVM is independent of the underlying blockchain protocol. This allows other blockchain implementations to adopt this standardized framework for contract execution.\newline
\textit{Polygon} is a so-called sidechain and aims to address the scalability limitations of Ethereum. The adoption of the EVM means that both blockchains share the same user address schemes. It further enables developers to deploy and execute Ethereum-based smart contracts on the Polygon network with minimal changes. \citep{matic_whitepaper}
%A high-level programming language is typically used to write smart contracts, and a compiler to convert them into bytecode \citep{mastering_ethereum}. Solidity is currently the dominant smart contract language.
%Polygon PoS is a so-called sidechain, closely associated with Ethereum mainnet and used for scaling.
%Ethereum Mainnet is the chain on which much of the Decentraland infrastructure has been deployed.
%In addition, \cite{eip1014} introduces a new contract creation mechanism allowing for more predictable contract addresses.%EIP-1014
%This allows developers to calculate the address of a contract before it is deployed.

\subsection{Tokens and Token Standards}
\cite{roth2019tokenization} define \textit{tokens} as rivalrous, digital units of value that represent ownership of an asset or utility. Smart contracts are the primary method of creating tokens on EVM-based blockchains. In essence, a smart-contract-based token is a mapping of accounts with token balances and a set of functions defining how these balances can be changed. Any smart contract containing these elements can be interpreted as a token contract. \citep{roth2019tokenization} \newline
\textit{Token standards} specify basic interfaces that allow for interoperability between contracts. These standards do not prescribe an implementation but set minimum requirements without restricting the design beyond that \citep{mastering_ethereum}. 
The most common token standards are ERC-20\footnote{https://eips.ethereum.org/EIPS/eip-20} for fungible tokens, ERC-721\footnote{https://eips.ethereum.org/EIPS/eip-721} for non-fungible tokens (NFTs) and ERC-1155\footnote{https://eips.ethereum.org/EIPS/eip-1155} for semi-fungible tokens.\newline
When a transaction completes, it produces a transaction receipt  that contains log entries providing information about the actions that occurred during its execution. Solidity\footnote{Solidity is the dominant high-level programming language for smart contracts targeting the EVM.} high-level objects called ``Events`` construct these logs, which can be queried from a full node and are stored separately from the state. \citep{mastering_ethereum} \newline
The standarization of tokens allows us to listen for ``transfer events``, which are emitted whenever an token changes ownership.
These transfer events include the \textit{sender} and \textit{recipient} address, along with a \textit{token ID} and/or \textit{amount}.

@@

\subsection{Decentraland}
\textit{Decentraland} is the first large-scale blockchain-based virtual world. Well-known companies like Tommy Hilfiger, Samsung, PepsiCo, Diesel, Adidas, and Netflix, who are actively engaging in Decentraland  \citep{metaverse-retailing2023}.\newline
Decentraland is an ideal platform for empirical research for two main reasons.
\textit{First}, its open architecture allows compiling snapshots of user activity, including the users' locations  within the metaverse. This was done by \cite{metaverse-retailing2023}, who captured snapshots of avatar locations over a period of 10 months. The addresses they collected are used for this thesis.
\textit{Second}, users who connect their Web3 account can own various digital assets, such as monetary units, land parcels, and avatar collectibles (e.g. wearables, emotes, and names). Smart contracts on Ethereum and Polygon track and manage the ownership of these assets \citep{goldbergschaer2023}. This allows us to analyze the entire transaction history and derive information about the users' economic activity. %avatars'
\newline
Initially, Decentraland launched with two native tokens. \textit{LAND}, an NFT compliant with ERC-721, manages the ownership of land parcels in Decentraland. \textit{MANA}, a fungible token compliant with ERC-20, is the in-world currency used to purchase digital goods and services. The majority of purchases and trades are settled in MANA. \newline 
Moreover, the Decentraland core developers provide a smart-contract-based marketplace where users may buy or exchange LAND or other in-game collectibles. %developers/core team

@@

\subsection{Non-custodial crypto asset mixers}
According to \cite{nadler2023tornado}, crypto-asset mixers are currently the most widely used approach to achieve privacy on public blockchains. At a high level, we can distinguish between custodial and non-custodial crypto-asset mixers. \newline
In the custodial setup, users send their assets to a centralized service provider's public deposit address. After some delay, the provider sends the assets back to a privately relayed recipient address. This approach is fully trust-based, as the service provider has complete control over the assets and access to the identifying data. \newline 
In contrast, non-custodial crypto asset mixers (e.g., Tornado Cash) replace the trusted mixing party with a publicly verifiable smart contract. They rely on cryptographic schemes that allow anyone to prove and independently verify the validity of a withdrawal without disclosing the link to a specific deposit. This has the advantage that users do not have to share the identifying information with anyone and there is no liquidity risk as the funds are locked. \citep{nadler2023tornado} \newline
In a nutshell, the mixer works in the following way: Various entities deposit the same amount of a specific crpyto asset to a mixer address acting as a pool. Anyone who has contributed to the pool may then generate a new address and withdraw their funds without revealing the link between the deposit and withdrawal addresses. This is achieved using zkSNARKs (zero knowledge, succinct, non-interactive argument of knowledge). 
Each depositor inserts a hash value in a Merkle-tree. At the time of withdrawal, each legitimate withdrawer can prove unlinkably with a zero-knowledge proof that they know the pre-image of a previously inserted hash leaf in the Merkle-tree. %Subsequently, users can withdraw their asset from the mixer whenever they consider that the size of the anonymity set is satisfactory.
 For a more detailed description of how Tornado Cash works, see \cite{nadler2023tornado} or \cite{Beres2020}. \newline
In the context of this work, it is sufficient to know that third parties can still observe the addresses that have deposited to and withdrawn from the pool. The set of all deposits that a particular withdrawal could be originated from is defined as the \textit{anonymity set}. 
In theory, a larger anonymity set will provide a higher level of privacy, since  any withdrawal address could be linked to more deposit addresses. However, in practice this does not always hold true. %this is not always the case
Some users reveal the link between their withdrawal and deposit addresses, which also affects the privacy of other depositors. For example, by withdrawing from the pool with the same address they deposited or directly send funds between the two addresses. According to \cite{nadler2023tornado}, careless use of the protocol, testing transactions or external incentive mechanisms might be reasons for this behaviour. \newline
In addition, \cite{Beres2020} and \cite{wu2022tutela} propose approaches to de-anonymize Tornado cash transactions using various clustering heuristics. These methods are discussed in more detail in section 4.

@@

\section{Data Collection and Preparation}
% Introduction Satz - Ausgangslage 
The starting point for our work was a dataset comprising 473,927 distinct (Web3) addresses. %The original dataset comprised 473,927 distinct (Web3) addresses.
 As we focus on clustering methods using on-chain data, we exclude addresses that have not been recorded on either Ethereum or Polygon. To accomplish this, we used data from Blockscan\footnote{\url{https://blockscan.com/}}. Overall, 59,651 addresses were recorded on Ethereum, 129,988 on Polygon, with 52,095 addresses appearing on both networks simultaneously. The set of ``clusterable`` addresses therefore counts 137,544 addresses. Subsets are visualized as a Venn diagram in Figure \ref{fig:Venn}.

@@

\subsection{Transactions and Token Transfer Events}
The address subsets were used to collect transaction and token transfer event data from Ethereum and Polygon. Ethereum or Polygon data can be accessed directly through a node or an application programming interface (API) provider like Infura or Etherscan. The absence of account indexing in Ethereum or Polygon poses a challenge for retrieving all past transactions and token transfers of a specific address, as it requires scanning through all blocks and token transfer events emitted by designated token contracts. Fortunately, Etherscan offers an API Endpoint Module for ``Accounts`` that facilitates the retrieval of transactions and token transfer events for a given address. By using our network-specific address subsets, we were able to significantly decrease the number of required API calls. 
We gathered all transactions and token transfers up until block \texttt{17,670,000} on Ethereum (July 11, 2023) and block \texttt{44,990,000} on Polygon (July 12, 2023). In total, we collected more than 30 million %30,689,978
token transfer events and more than 16 million %16,092,531
 normal transactions. While the amount of transactions is similar between the two networks, there are three times more transfer events on Polygon. \newline The output was stored in separate CSV files and also imported into a MongoDB\footnote{\url{https://www.mongodb.com/}} database.The database contains distinct collections for transactions and transfers. For information on the different data fields, see Appendix.

@@

\subsection{Intra-set asset transfers}
In this project, we want to use entity clustering to determine the number of entities that are represented by the remaining 137,544 addresses. Or, to phrase it differently, do users interact with Decentraland using multiple EOAs, and can we identify addresses that belong to the same user?
\newline
To achieve this, we often only need to consider transactions or token transfer events where both the ``\texttt{from}`` (sender) and the ``\texttt{to}`` (recipient) address are within the address set. Regarding normal transactions, this means that we only keep native asset transfers (Ether or Matic) as almost all addresses are EOAs. \newline
The intra-set asset transfer dataset is mainly used to generate the network graphs (see section xx). When clustering addresses using graph-based approaches, this has the advantage that we do not need to differentiate between EOAs and smart contract account later on. Furthermore, it allows us to observe and visualize the asset flows between users. In total, we have 974,478 intra-set token transfers and 210,548 intra-set native asset transfers (101,621 Matic, 108,927 Ether).

\iffalse
%As nearly all addresses are Externally Owned Accounts (EOAs), we do not need to differentiate between EOAs and smart contract accounts, which further simplifies the clustering process. \newline
%(logging in to Decentraland with multiple addresses) When it comes to entity clustering, we essentially want to answer the question of how many real-world entities our remaining  137,544 addresses represent. 
-Also for other clustering heuristics we often make this reduction, for example: In the self-approval approach, we only consider approval transactions where another address from the set is approved.
-filter our transactions and token transfers to entries where both the 'from' and the 'to' are within this narrowed pool of addresses. We define this as intra-set asset transfers. For the transactions, this means that we only consider native asset (Ether or matic) transfers. %between two EOAs
-As already mentioned in the introduction, the goal of this paper is to cluster addresses within who (1) were active in Decentraland during the 9-month period (e.g., within the initial dataset) and (2) transacted or interacted with any token on Ethereum or Polygon. 
-Essentialy, we want to answer two main research questions when it comes to clustering:
1. How many unique entities does the address set represent (i.e. were active on Decentraland during the 9-month period)?
2. Are users interacting through multiple accounts/EOAs/addresses with Decentraland?
-we focus on on-chain data (and not user location or avatar data), allowing us to exclude addresses that have not transacted or interacted with any token on Ethereum or Polygon. 
-As mentioned in the introduction, the goal of this paper is to cluster addresses belongig to the same entity that were recorded/active in Decentraland during the 9-month period. Therefore, we disregard addresses outside of this set. Also, since we only look at on-chain data, we cannot cluster addresses without activity on Polygon or Ethereum. Doing these two reductions, there are  141, 591 'clusterable' addresses remaining.
-Since all of our addresses are EOAs (some multisig smart contract wallets). -> Check this?
\fi

\subsection{ENS ground-truth pairs}
Similarly to \cite{Beres2020}, we use Ethereum Name Service (ENS) identifiers as ground truth information to evaluate some of the clustering methods.\newline
\textit{ENS} is a naming system that relies on the Ethereum smart contracts. Its main purpose is to map human-readable names, such as \texttt{`alice.eth`}, to machine-readable identifiers, mostly Ethereum addresses. The architecture of ENS comprises two main components: the registry and resolvers. \newline
The \textit{ENS registry} consists of a single smart contract\footnote{The registry contract is ERC-721 compliant, which means that .eth registrations can be transferred in the same way as other NFTs.} that maintains a list of all domains and subdomains, recording the ``owner`` and ``resolver`` for each. The registry allows the owner of a domain to make changes to that data. \citep{ENSdocs} %Domain owners may set the resolver for the domain.
\newline
\textit{Resolvers} are responsible for %(the actual process of)
 translating names into addresses. Any contract that adheres to the relevant standards may act as a resolver. The process of resolving a name involves two steps: \textit{First}, the registry must be queried to find out which resolver is responsible for the name, and \textit{second}, that resolver must be asked for the response to the query. \newline
In addition to regular resolution from name to address, ENS also supports ``reverse resolution``, allowing for a mapping from address back to a name (or other metadata). Reverse resolution is accomplished via %special
 domain \texttt{`addr.reverse`}% and the resolver function \texttt{name()}.
. This domain is owned by a special purpose registrar contract that allocates subdomains to the owner of the matching address - for instance, the address \texttt{`0x1234\dots`} may claim the name \texttt{`1234\dots.addr.reverse`}, and configure a resolver and records on it. The resolver in turn supports the \texttt{name()} function, which returns the name associated with that address.\newline
 Reverse resolution %(without a library)
 follows the same pattern as forward resolution: Get the resolver for \texttt{`1234....addr.reverse'} (where \texttt{`1234...`} is the address you want to reverse-resolve), and call the \texttt{name()} function on that resolver. However, ENS does not enforce the accuracy of reverse records. This means \texttt{`1234....addr.reverse`} may falsely claim that the name associated to their address is \texttt{`alice.eth`}.
